Lossy compression of DNA sequencing quality data

### 2015-02-04 manuscript/q-compression.tex
Continuing the revision process. Consolidating information in the Methods section, including text that was previously elsewhere. I have changed, slightly, our definition of MSE to be consistent with the standard definition.

Sequencing read data constitutes the largest category of sequence data
in genomics research because of great redundancy, inclusion of quality
values, and read-level naming and metadata. Because of this abundance
effective compression of read data has the potential for substantial
improvement in data storage and transfer efficiency.  Among the
several challenges for compression of read data is dealing with
different error profiles resulting from differences in underlying
chemistries, signal detection and processing mechanisms, inherent
biases, and other idiosyncratic properties of distinct high-throughput
sequencing technologies. Illumina sequence reads, the focus of this
research, are characterized by having relatively few insertion and
deletion errors, but substitution (miscall) errors are much more
frequent and have context-specific patterns. These errors are
non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low quality tails at the 3$^{\prime}$
end~\cite{Minoche:2011km}).

@@

Quality values are typically used throughout bioinformatics pipelines.
Software such as \textsc{fastqc}~\cite{citeulike:11583827} and
Prinseq~\cite{Schmieder:2011gd} provide the user with detailed
descriptions of the distribution of sequence quality scores, and these
scores can be used to guide filtering and trimming of sequence read
data as processing step before subsequent analyses. Quality values can
be used by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Gnerre:2011kx}.  Short-read alignment
software, such as Bowtie 2~\cite{Langmead:2012rw}, use quality values
to scale the mismatch penalty in a way that mismatches between poor
quality bases of the query and the reference are weighted less than
mismatches occurring on high quality bases. Similar to alignment,
software for detecting single nucleotide polymorphisms (\textsc{snp}s)
can use quality values~\cite[e.g.,][]{McKenna:2010bh}, and identified
\textsc{snp}s with high-quality calls are deemed more reliable than
those with low-quality calls, particularly in low coverage regions.

Previous literature on sequence data compression has largely focused
on lossless compression of base calls~\cite[reviewed
  in][]{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
Although some recent work has included or focused on compression of
quality
values~\cite[e.g.,][]{Tembe:2010ys,Kozanitis:2011kl,asnani2012lossy,Hach:2012ys,Wan:2012kq,janin2013adaptive,Ochoa:2013rt,Canovas:2014fr,zhou2014compression,DBLP:conf/recomb/YuYB14}.
Typical compression methods fail to utilize the biases inherent in
existing sequencing technologies. Depending on the sequencing
technology used, the base calling is more accurate in the earlier
cycles, gradually losing quality in later
cycles~\cite{Kozanitis:2011kl}. To this end, methods for compressing
these quality sequences utilize these general quality
trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
that builds a fixed-state Markov model on adjacent quality scores and
uses Huffman Coding for compression based on the transition
probabilities.

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective some absolute loss of
information is deemed acceptable if the inferences from analyses are
relatively uneffected. Therefore we here describe our research
investigating some general strategies for compression of sequence read
quality values. Our objective provide perspective on several
strategies logically applied to Illumina sequence read quality values
rather than to generate a robust high-quality software for
use. Consistent with our perspective we place emphasis not only on the
compression efficiency, but also the effect of lossy compression on
typical subsequent analyses.  We examine three general classes of
lossy compression methods -- binning, modeling, and profiling -- and
consider an exemplar of each class.

@@

\subsection{Evaluating lossy compression algorithms}

@@

For any lossy compression method, we need to evaluate the tradeoff
between its physical compression and the loss in its accuracy. Here,
we use mean bits/base-call to measure the physical compression of the
data and mean squared error (\textsc{mse}) to measure the loss in
quality. Bits/base-call is calculated simply as the compressed file
size (in bits) divided by the total number of quality
values. \textsc{mse} is calculated as
$\frac{1}{n}\sum_{i=1}^{n}{(Q_i-Q_i')^2}$, where $n$ is the number of
sequences, $Q_i$ is the original quality value associated with
sequence position $i$, and $Q_i'$ is the corresponding compressed
quality value.

>>>>

Read data from high-throughput sequencing constitutes the largest
category of data in genomics research because of great redundancy,
inclusion of quality values, and read-level naming and
metadata. Because of this abundance effective compression of read data
has the potential for substantial improvement in data storage and
transfer efficiency. Among the several challenges for compression of
read data is dealing with different error profiles resulting from
differences in underlying chemistries, signal detection and processing
mechanisms, inherent biases, and other idiosyncratic properties of
distinct high-throughput sequencing technologies. Sequence reads
generated by instruments such as an Illumina HiSeq, the focus of this
research, are characterized by having relatively few insertion and
deletion errors, but substitution (miscall) errors are much more
frequent and have context-specific patterns. These errors are
non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low quality tails at the 3$^{\prime}$
end~\cite{Minoche:2011km}).

@@

Quality values are typically used throughout bioinformatics pipelines.
Software such as \textsc{fastqc}~\cite{citeulike:11583827} and
Prinseq~\cite{Schmieder:2011gd} provide the user with detailed
descriptions of the distribution of sequence quality scores, and these
scores can be used to guide filtering and trimming of sequence read
data as processing steps before subsequent analyses. Quality values
can be used by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment software, such as Bowtie 2~\cite{Langmead:2012rw}, use
quality values to weight mismatches between query and reference
sequences.  Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values, and identified \textsc{snp}s
with high-quality calls are deemed more reliable than those with
low-quality calls, particularly in low coverage
regions~\cite[e.g.,][]{McKenna:2010bh}.

Previous literature on sequence data compression has largely focused
on lossless compression of base calls~\cite[reviewed
  in][]{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
Although some recent work has included, or focused on, compression of
quality
values~\cite[e.g.,][]{Tembe:2010ys,Kozanitis:2011kl,asnani2012lossy,Hach:2012ys,Wan:2012kq,janin2013adaptive,Ochoa:2013rt,Canovas:2014fr,zhou2014compression,DBLP:conf/recomb/YuYB14}.
Typical compression methods fail to utilize the biases inherent in
existing sequencing technologies. Depending on the sequencing
technology used, the base calling is more accurate in the earlier
cycles, gradually losing quality in later
cycles~\cite{Kozanitis:2011kl}. To this end, methods for compressing
these quality sequences utilize these general quality
trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
that builds a fixed-state Markov model on adjacent quality scores and
uses Huffman Coding for compression based on the transition
probabilities.

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective some absolute loss of
information is deemed acceptable if the inferences from analyses are
relatively unaffected. Therefore we here describe our research
investigating some general strategies for compression of sequence read
quality values. Our objective provide perspective on several
strategies logically applied to Illumina sequence read quality values
rather than to generate a robust high-quality software for
use. Consistent with our perspective we place emphasis not only on the
compression efficiency, but also the effect of lossy compression on
typical subsequent analyses.  We examine three general classes of
lossy compression methods -- binning, modeling, and profiling -- and
consider an exemplar of each class.

@@

\subsection{Performance evaluation}

As a measure of compression effectiveness we use bits/base-call, and
define it as the size of the compressed representation of quality
values (in bits) divided by the number of quality values represented.
As a measure of information loss we use mean squared error
(\textsc{mse}) as aloss function, and define it as
$\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
sequences, $Q_i$ is the original quality value associated with
sequence position $i$, and $Q_i'$ is the corresponding
compressed/decompressed quality value.

@@
